\documentclass[sigconf]{acmart}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{multirow}
\usepackage[dvipsnames]{xcolor}
\usepackage{svg}
\usepackage{amsmath}
\usepackage{makecell}
\usepackage{pifont}
\newcommand{\xmark}{\ding{55}}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric, fit, backgrounds, calc}
\DeclareMathOperator*{\argmax}{argmax}
\title{
Incremental Tree Ensembles for Evolving Feature Spaces Under Structured Missingness
}

\author{Anonymous Author(s)}
\affiliation{\institution{Under Double-Blind Review} \country{Unknown}}
\email{anonymous@example.com}

\begin{document}

\begin{abstract}
Machine learning models deployed in healthcare, finance, and other dynamic environments face a persistent challenge: feature spaces that evolve over time as new measurements, tests, or data sources become available. This creates \textit{structured missingness} - newer features are systematically absent from historical records, requiring models to serve heterogeneous subpopulations with different feature availability. Traditional approaches require full retraining or maintaining multiple specialized models, both costly and incompatible with privacy-constrained settings where data cannot be centralized.

We introduce an incremental learning framework for Gradient Boosting Machines that seamlessly integrates new features into existing models without full retraining. Our two-phase approach trains a base model on commonly available features, then warm-starts an extended model that incorporates newly available features through adaptive tree pruning. Crucially, the phases operate independently - only the model artifact is transferred, not raw data - enabling deployment under privacy constraints and evolving feature availability.

Evaluation across 10 diverse datasets demonstrates that incremental learning without data sharing matches standard training: per-dataset AUCs are within $\sim 1\%$ of a data-sharing combined model on every dataset and the population-weighted AUC gap is indistinguishable from zero in expectation. In addition, we benchmark against five baselines for learning under evolving feature spaces - AdaptiveXGBoost, PUFE, OCDS, EMLI, and GBDT-IL - under an identical preprocessing and data-split protocol; our framework's mean population-weighted AUC is $0.04$ to $0.14$ above each of them.
\end{abstract}

\keywords{Incremental Learning, Structured Missingness, Gradient Boosting, Feature Evolution, Adaptive Tree Pruning, Privacy-Preserving ML, Model Transfer}

\maketitle

\section{Introduction}

Machine learning models are increasingly deployed in dynamic environments where the feature space evolves over time. In healthcare, patients may undergo new diagnostic tests during hospitalization, leading to heterogeneous feature availability across records \cite{rajkomar2018healthcare}. In finance, regulatory changes and behavioral metrics are introduced incrementally and collected only for certain user segments \cite{dixon2020finance}. These evolving schemas lead to structured, non-random missingness where new features are systematically absent from earlier records not due to noise, but as a result of how data is acquired. Importantly, these features do not update at uniform intervals, creating temporal asymmetry in data availability.

Tree-based models, particularly GBMs, remain the dominant choice in these domains due to their strong performance on tabular data \cite{grinsztajn2022why,rana2023comparative,langenberger2023hcp}, natural handling of mixed data types and missing values, and inherent explainability - a critical requirement in healthcare and finance applications.

Maintaining robust predictive models in such settings presents a major challenge. Organizations often resort to training multiple specialized models aligned with different feature subsets. However, this strategy introduces substantial maintenance burden, complicates deployment workflows, and risks version drift across models. Moreover, it is incompatible with real-world scenarios involving partial feature availability at inference time, particularly under privacy constraints such as GDPR where feature usage may vary at the user level.

A closely related challenge arises in cross-organizational collaboration, where multiple institutions hold complementary feature sets but cannot share raw data due to privacy regulations. For example, two hospitals may collect different diagnostic measurements for overlapping patient populations, or banks collaborating on anti-money laundering (AML) models may hold transaction features that cannot be exchanged. Our framework directly addresses this scenario: one institution trains a base model on its features and shares only the model artifact, while the other extends it with private features using warm-start and adaptive pruning. Our experimental design mirrors exactly this setup - the base model is trained on one feature set and extended with additional features from a different subpopulation, with no data shared between phases.

\paragraph{Contributions.}
This work addresses the problem of learning under evolving feature spaces with structured missingness in settings where raw data cannot be shared across training phases. Our contributions are fourfold:
\begin{itemize}
\setlength\itemindent{0pt}
\setlength{\itemindent}{0pt}
\item \textbf{(1) A formalisation of column-wise incremental learning under structured missingness and data-separation constraints.}
We introduce a learning setting in which new features become available over time or across institutions, while historical data lacks these features in a systematic (non-random) manner. We formalise this as a two-phase learning problem with a hard data-separation constraint, and define a population-aware objective that jointly evaluates performance on feature-absent and feature-present subpopulations. This formulation captures practical deployment requirements in domains such as healthcare and finance, where both feature availability and data sharing are inherently constrained.

\item \textbf{(2) A population-aware incremental learning framework for gradient boosting models.}
We propose a two-phase framework that extends an existing gradient boosting model with newly available features without requiring access to the original training data. The approach combines warm-start initialisation, validation-driven adaptive tree pruning, and continued boosting to enable controlled knowledge transfer between phases. The resulting model maintains a single deployable predictor that dynamically adapts to heterogeneous feature availability at inference time, while preserving performance guarantees on legacy data.

\item \textbf{(3) Empirical evidence that incremental learning without data sharing achieves parity with full-data training.}
Through a comprehensive evaluation on 10 real-world datasets with both natural and controlled structured missingness, we show that the proposed framework consistently matches the performance of a data-sharing baseline trained on all features jointly. Across datasets, the population-weighted AUC difference is statistically indistinguishable from zero (paired Wilcoxon $p = 0.121$ over $5\times 10 = 50$ seed-dataset pairs), while substantially outperforming existing methods for evolving feature spaces. These results demonstrate that, under structured missingness, population-level specialisation can compensate for the absence of joint training.

\item \textbf{(4) A unified experimental benchmark for evolving-feature learning.}
We provide a systematic comparison against five representative baselines spanning streaming learning, feature reconstruction, and metric learning approaches~\cite{montiel2020axgb,he2021pufe,he2019ocds,dong2021eml,chen2024gbdtil}, under a unified preprocessing and evaluation protocol. This benchmark isolates the effect of model design choices and highlights the importance of explicitly modelling heterogeneous feature availability.

\section{Related Work}
Missing data are traditionally addressed either by discarding incomplete instances or by imputing missing values, but each strategy has drawbacks~\cite{chen2016xgboost}. Common strategies include MICE~\cite{vanbuuren2011mice}, which iteratively imputes values using regression under MAR assumptions, and tree-based methods like CART~\cite{loh2011cart} and MIA~\cite{twala2008mia}, which handle missingness via surrogate splits or by treating missing values as informative categories.

Modern gradient boosting frameworks such as XGBoost~\cite{chen2016xgboost}, LightGBM, and CatBoost include built-in mechanisms for missing values, but these are fundamentally heuristic. XGBoost learns a default split direction at each tree node: when a feature is absent during inference, the sample is routed left or right based on whichever direction minimized training loss. This approach treats missingness as ``just another value'' and optimizes the routing empirically. Critically, it does not model \textit{why} a value is missing, does not distinguish between random and structured missingness, and does not adapt the model architecture for subpopulations with different feature availability. When missingness is structured - e.g., ICU lab tests ordered only for critical patients, or premium financial metrics collected only for high-value customers - the missing-value population and the observed-value population may exhibit fundamentally different predictive patterns. A single model with heuristic default directions cannot specialize for these distinct subgroups.

Bayesian Additive Regression Trees (BART)~\cite{chipman2010bart} can be augmented with MIA to natively support incomplete data without imputation, but similarly assume that missingness is random or independent. This limitation motivates our approach, which explicitly models structured missingness by training separate specialized models for each subpopulation rather than relying on heuristic missing-value routing within a single model.
Recent research has begun addressing more complex streaming and evolving-feature scenarios, where both the data distribution and feature space can change over time. While traditional stream learners assume a fixed set of input features, real-world systems often experience column-wise feature drift with new sensors, logs, or sources introducing features that were previously unavailable.
Several ensemble based methods, such as Adaptive XGBoost (AXGB)~\cite{montiel2020axgb}, adapt to row-wise concept drift by updating trees incrementally, often using sliding windows or tree replacement strategies. However, these models typically assume a static input space and cannot gracefully integrate new feature columns without full retraining. Other frameworks like Online learning from Capricious Data Streams (OCDS) \cite{he2019ocds} attempt to bridge feature sets by learning mappings between old and new features using generative models, but they still rely on conventional missing-value treatments and may struggle with structured missingness when the absence of data is systematic and correlated with the target. These limitations motivate our approach. Rather than retraining from scratch, we incrementally integrate new feature columns into an existing gradient boosting model, preserving previously learned structure and enabling adaptation to evolving schemas and feature sets - even as data distributions shift.

ORF3V is an online forest-based model designed to handle data streams with changing feature spaces~\cite{beyazit2019orf3v}. Each feature is tracked by its own mini-ensemble of shallow trees, allowing the model to add or remove features dynamically. Unlike our method, ORF3V maintains a separate model per feature and does not distinguish between base and extended phases. Our approach focuses on structured feature allocation and controlled model updates for deployment simplicity.
Adaptive XGBoost (AXGB)\cite{montiel2020axgb} updates an ensemble online by adding/replacing trees on mini-batches and triggering concept-drift detectors but assumes a fixed feature space, while Streaming Gradient Boosted Trees (SGBT)\cite{vaquet2022sgbt} embeds a tree-replacement mechanism for tracking distribution shifts in data streams with unchanged features; in contrast, our method tackles the complementary problem with a static population but growing columns, grafting additional features into the booster through extra boosting rounds and targeted pruning.
GBDT-IL is an incremental boosting method that updates the model over time by adding new trees and removing outdated ones~\cite{chen2024gbdtil}. It selects useful features using the Fisher Score, but does not handle situations where some features are missing. In contrast, our method is designed for evolving feature sets and can adapt when only part of the features are available.
PUFE~\cite{he2021pufe} and Evolving Metric Learning (EML)~\cite{dong2021eml} both deal with changing feature spaces over time. PUFE fills missing overlaps using matrix completion and combines predictions from multiple models in an ensemble, while EML transforms vanished features into a compact form and integrates new ones using a learned distance metric in two stages. In contrast to these approaches, our method focuses on structured missingness in static tabular data where new features are added rather than vanishing. Our method extends a base GBM with new features in a second phase, preserving the original model without retraining or ensembling.
LIR-eGB. LIR-eGB~\cite{wang2023liregb} is designed for data streams where the distribution changes over time (concept drift). It improves a GBM by pruning weak trees that no longer help, based on a loss improvement ratio (LIR). The model switches between naive and statistical pruning, and adds new trees incrementally to adapt to the changing data. In contrast, our method focuses on static tabular data with structured missing features. We extend an existing GBM by adding new features in a second phase while preserving useful base trees, without retraining or handling concept drift.
\begin{table}[ht]
  \caption{Capability matrix (rows = method, columns = key capability)}
  \label{tab:method-comparison-transposed}
  \centering
  \resizebox{0.95\columnwidth}{!}{%
  \begin{tabular}{l c c c c c}
    \toprule
    & \makecell[c]{\textbf{Column-wise}\\\textbf{incremental}\\\textbf{learning}}
    & \makecell[c]{\textbf{Persistent}\\\textbf{feature}\\\textbf{support}}
    & \makecell[c]{\textbf{Handles}\\\textbf{structured}\\\textbf{missingness}}
    & \makecell[c]{\textbf{Single}\\\textbf{unified}\\\textbf{model}}
    & \makecell[c]{\textbf{Code}\\\textbf{Provided}}\\
    \midrule
    \textbf{AXGB} & $\times$ & $\times$ & $\times$ &  \checkmark & \checkmark \\
    \textbf{SGBT} & $\times$ & $\times$ & $\times$ & \checkmark & \checkmark \\
    \textbf{OCDS} & \checkmark & \checkmark & \checkmark & $\times$ & $\times$ \\
    \textbf{ORF3V} & \checkmark & \checkmark & $\times$ & $\times$ & $\times$\\
    \textbf{GBDT-IL} & $\times$ & \checkmark & $\times$ & \checkmark & $\times$ \\
    \textbf{PUFE} & \checkmark & $\times$ & \checkmark &  \checkmark & $\times$ \\
    \textbf{LIR-eGB} & $\times$ & $\times$ & $\times$ & \checkmark & \checkmark \\
    \textbf{EML}  & \checkmark & $\times$ & \checkmark &  \checkmark & $\times$ \\
    \textbf{Our}         &  \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
    \bottomrule
  \end{tabular}%
  }
\end{table}
\section{Problem Formulation}
\label{sec:problem}

We formalise the setting in which column-wise incremental learning is required and specify the objectives that our method must meet.

\subsection{Data Model and Notation}

Let
\[
\mathcal D=\{(\mathbf{x}_i,y_i)\}_{i=1}^{N},
\qquad
y_i\in\{0,1\},
\]
be a binary-classification dataset collected over a time horizon
\(0<t_1<t_2<\dots<t_N\).
Each input vector \(\mathbf{x}_i\in\mathbb R^{p_B+p_E}\) is partitioned into
\[
\mathbf{x}_i=\bigl[\;\mathbf{b}_i\;\|\;\mathbf{e}_i\bigr],
\qquad
\mathbf{b}_i\in\mathbb R^{p_B},\;
\mathbf{e}_i\in\mathbb R^{p_E},
\]
where \(\mathbf{b}_i\) is the base-feature block that is always observed, and
\(\mathbf{e}_i\) is the extended-feature block, present only for instances that arrive
after a feature-release time \(t_E\) (or when acquisition conditions are met).

A binary missingness mask
\(\mathbf{m}_i\in\{0,1\}^{p_E}\) is defined by
\(m_{ij}=1\) if the \(j\)-th extended feature of instance \(i\) is observed and
\(m_{ij}=0\) otherwise.
We assume structured (NMAR) missingness:
\[
P(\mathbf{m}_i \mid \mathbf{b}_i,\mathbf{e}_i,y_i)
\;\neq\;
P(\mathbf{m}_i \mid \mathbf{e}_i,y_i),
\]
i.e.\ the probability that a value is missing may depend on observed base features (
In healthcare, advanced laboratory tests are ordered only when certain vitals exceed thresholds; in finance, premium behavioural metrics are gathered only for high-value customers).

\subsection{Predictive Objective}

We seek a single probabilistic classifier
\[
f_\theta:\;
  \mathbb R^{p_B+p_E}\times\{0,1\}^{p_E}
  \;\longrightarrow\;[0,1],
\]
parameterised by \(\theta\), that satisfies:

\begin{enumerate}[label=(\alph*)]
\item \textbf{Validity under partial information.}\;
For any instance \(i\), inference uses only the available features:
\[
\hat y_i
  = f_\theta\!\Bigl(
      \bigl[\mathbf{b}_i \,\|\, \mathbf{m}_i\odot\mathbf{e}_i\bigr],
      \mathbf{m}_i
    \Bigr),
\]
where "\(\odot\)" denotes element-wise multiplication (missing entries are zeroed).

\item \textbf{Minimisation of predictive error.}\;
The optimisation target is the binary log-loss evaluated on a validation set:
\[
\mathcal L_{\mathrm{pred}}(\theta)=
\frac{1}{|\mathcal D_{\mathrm{val}}|}
\sum_{(\mathbf{x}_i,y_i)\in\mathcal D_{\mathrm{val}}}
\Bigl[
  -y_i\log\!\bigl(\hat y_i\bigr)
  -(1-y_i)\log\!\bigl(1-\hat y_i\bigr)
\Bigr],
\]
with \(\hat y_i\) defined as above.
\end{enumerate}

\subsection{Incremental-Learning Constraint}

Model training proceeds in two phases.  Let \(\theta^{(0)}\) denote the parameters of
a base model trained solely on \(\{\mathbf{b}_i\}\).  Extending this model
yields
\begin{equation}
\label{eq:inc-opt}
\theta^{\star}
=
\arg\!\min_{\theta}
\;
\underbrace{\mathcal L_{\mathrm{pred}}(\theta)}_{\text{predictive performance}}
\;+\;
\lambda\,\underbrace{C_{\mathrm{inc}}\!\bigl(\theta\mid\theta^{(0)}\bigr)}_{\text{incremental cost}}.
\end{equation}
where \(C_{\mathrm{inc}}(\theta\mid\theta^{(0)})\) measures the additional computation
required to transform \(\theta^{(0)}\) into the extended-feature model \(\theta\),
and \(\lambda>0\) trades off predictive gain against update cost.

\paragraph{Deployment guarantees.}
For optimisation~\eqref{eq:inc-opt} to be acceptable in production we require:

\begin{enumerate}[label=(\Roman*)]
\item No degradation on legacy traffic:
\[
\operatorname{AUC}\!\bigl(f_{\theta^{\star}};\{\mathbf{m}_i=\mathbf 0\}\bigr)
\;\ge\;
\operatorname{AUC}\!\bigl(f_{\theta^{(0)}};\{\mathbf{m}_i=\mathbf 0\}\bigr).
\]

\item \emph{Monotonic benefit on enriched traffic:}
for the subset with any \(m_{ij}=1\),
\[
\operatorname{AUC}\!\bigl(f_{\theta^{\star}}\bigr)
\;\ge\;
\operatorname{AUC}\!\bigl(f_{\theta^{(0)}}\bigr).
\]
\end{enumerate}

Given \((\mathcal D,\mathbf m)\) and trade-off \(\lambda\), our goal is to design an
\textbf{incremental training procedure} that solves~\eqref{eq:inc-opt} whenever new
columns appear, thereby delivering a single deployable model
\(f_{\theta^{\star}}\) that: (i) attains near-optimal AUC on both basic and
extended cohorts, (ii) controls incremental cost through adaptive pruning and
continued boosting, and (iii) gracefully handles heterogeneous features
availability.

\section{Methods}
\label{sec:method}

Our approach implements a two-phase incremental learning framework for Gradient Boosting Machines, designed for evolving feature spaces with structured missingness (Figure~\ref{fig:framework}). Rather than retraining from scratch when new features become available, we incrementally extend an existing model through warm-starting and adaptive pruning. The framework consists of two components: \textbf{Phase~1} trains a base model on commonly available features; \textbf{Phase~2} transfers knowledge from the base model and extends it with newly available features. Critically, the two phases operate independently - only the model artifact is shared - enabling both incremental updates over time and privacy-constrained collaboration across institutions.

\subsection{Phase 1: Base Model Training}

First, we construct a base model using a gradient boosting machine (GBM) trained exclusively on commonly available features. This foundation model undergoes thorough hyperparameter tuning through Bayesian optimization to maximize performance on the initial feature set. The optimized hyperparameters include learning rate, tree depth, regularization parameters, and other model-specific configurations to ensure optimal validation set performance and generalizability.

The underlying predictive model utilizes gradient boosting machines (GBMs), an ensemble learning method that builds a series of decision trees sequentially, with each tree correcting errors made by previous trees. The base model is formulated as:

\begin{equation}
F_{base}(x) = \sum_{i=1}^{n} \alpha_i h_i(x_{F_{base}})
\end{equation}

Where $h_i$ are decision trees trained on the base features $F_{base}$, and $\alpha_i$ are tree weights.

\subsection{Phase 2: Extended Model Training}

The second phase develops an extended model that incorporates newly available features present in only a subset of instances, corresponding to Phase 2 in Figure~\ref{fig:framework}. Rather than training from scratch, we employ a transfer learning approach with three key mechanisms:

\begin{enumerate}
\item \textbf{Warm-start initialization} from the base model, preserving knowledge gained from widely available features

\item \textbf{Adaptive tree pruning} to determine the optimal subset of base trees to retain, strategically balancing prior knowledge with flexibility for new patterns

\item \textbf{Continued training} that allows the model to learn relationships involving the extended feature set
\end{enumerate}

The pruning level determining exactly how many trees $k$ to preserve from the base model is itself optimized through Bayesian optimization against a validation set. This process evaluates different pruning thresholds to identify the configuration that best transfers relevant knowledge while avoiding concept drift between the feature spaces.

The extended model builds upon this foundation through transfer learning:

\begin{equation}
F_{ext}(x) = \sum_{i=1}^{k} \alpha_i \, h_i(x) + \sum_{j=1}^{m} \beta_j \, g_j(x)
\end{equation}

where $\{h_1, \dots, h_k\} \subseteq \{h_1, \dots, h_n\}$ are the first $k\!\le\!n$ base trees retained after pruning (each still operating only on its trained base-feature splits), and $\{g_1, \dots, g_m\}$ are additional trees trained on the combined feature space $F_{base} \cup F_{ext}$. We write $h_i(x)$ and $g_j(x)$ in terms of the full input vector $x$ for compactness; each tree internally evaluates only the features it was trained on, so notation conflicts with~Eq.~1 (where the tree set is the unpruned $n$) are resolved by the explicit $i\in\{1,\dots,k\}$ summation index above.

\subsection{Feature Allocation in Practice and in Our Experiments}
\label{sec:feature-allocation}

\paragraph{Real-world scenarios.} The partition between $\mathcal{F}_{\text{base}}$ and $\mathcal{F}_{\text{ext}}$ is fixed by the deployment context, not chosen by the framework. Three canonical scenarios:
(i)~\emph{Evolving schema over time} - new sensors, lab tests, or data-collection systems come online; historical records carry only the legacy columns ($\mathcal{F}_{\text{base}}$), while newer records also carry the recently added columns ($\mathcal{F}_{\text{ext}}$).
(ii)~\emph{Cross-institutional collaboration without data sharing} - institution~A has trained a model on its columns ($\mathcal{F}_{\text{base}}$) and ships only the model artefact to institution~B, which then extends it with its private columns ($\mathcal{F}_{\text{ext}}$) on its own population.
(iii)~\emph{Per-customer feature availability} - a feature subset is observed only for some segments (premium users, consenting users under GDPR, hospitalised vs.\ outpatient cases), and the partition follows that boundary.
In each case the allocation is determined by data-source availability, regulation, or institutional ownership, so the framework does not search over it.

\paragraph{Allocation in our benchmark.} Public benchmarks do not expose a source-derived split, so we cannot reuse the deployment-context boundary in our evaluation. Instead, we treat allocation selection as a \emph{benchmark-construction} step that is external to the framework itself. For each dataset we restrict the candidate pool to allocations consistent with structured missingness --- natural null groups for the natural datasets and pre-specified conditional null-injection patterns (confirmed NMAR by the validation in Section~\ref{sec:missingness-validation}) for the augmented datasets --- screen the resulting candidate combos, and fix the representative allocation that exhibits the base/extended partition the framework is designed for. The chosen allocation is then frozen for every subsequent comparison; baselines, ablations, and seed-sweep runs all consume the same per-dataset $(\mathcal{F}_{\text{base}}, \mathcal{F}_{\text{ext}})$ split listed in Table~\ref{tab:dataset-summary}. Bayesian optimisation~\cite{optuna} inside each evaluation tunes only XGBoost hyperparameters and the Phase~2 pruning level $k$, never the allocation. We are explicit about this because the screening step is required to construct the benchmark but is \emph{not} part of the framework's deployment loop: a real-world user receives $\mathcal{F}_{\text{base}}$ and $\mathcal{F}_{\text{ext}}$ from their data sources and runs Phases~1--2 directly, with no allocation search. Generality of our parity claim is therefore conditional on a deployment whose natural allocation has the same NMAR structure as the screened benchmark allocations; we view this as the regime the framework targets, and as the same regime captured by the missingness-validation analysis.

\subsection{Inference}

The inference process, depicted in the lower section of Figure~\ref{fig:framework}, demonstrates how new data is routed to either the base or extended model depending on feature availability. During inference, the system automatically determines which model configuration to use based on the available features for each instance:

\begin{itemize}
\item \textbf{Base model inference}: Applied when only base features are available
\item \textbf{Extended model inference}: Applied when both base and extended features are present
\item \textbf{Adaptive routing}: The system seamlessly switches between models based on real-time feature availability
\end{itemize}

This comprehensive approach enables ongoing model enhancement as new features emerge, without requiring complete retraining or sacrificing performance on instances with limited feature availability.

\subsection{Performance Floor via the $k=0$ Fallback}
\label{sec:k0-fallback}

The pruning step in Phase~2 is a search over how many base trees to retain, $k \in \{0, 1, \ldots, |T_{\text{base}}|\}$. We do not claim a formal theoretical bound; instead we note a simple \emph{design property}: the search space includes $k=0$, which corresponds to discarding the base model entirely and training the extended model from scratch on $\mathcal{F}_{\text{base}}\cup\mathcal{F}_{\text{ext}}$. Because $k$ is selected by validation performance, whenever no positive $k$ improves over the from-scratch alternative the search returns the $k=0$ branch.

In practice this gives the framework a built-in floor: when the base model transfers useful structure the search keeps a positive $k$, and when it does not the framework reduces to training from scratch on the extended-feature subset --- the same model the data-sharing baseline would train if it were restricted to the has-extended population. This is a procedural property of the search procedure, not a worst-case theorem.

\begin{algorithm}[t]
\caption{Adaptive Tree Pruning for Incremental Learning in Gradient Boosting}
\begin{algorithmic}[1]
\Require Base model $M_{base}$ with trees $T_{base} = \{t_1,...,t_n\}$
\Require New data batch $D_{new}$ with features $F$
\Ensure Updated model $M_{updated}$
\Function{FindOptimalPruning}{$M_{base}$, $D_{new}$, $n_\text{trials}$}
    \State $D_{train}, D_{val} \gets$ SplitData($D_{new}$)
    \State $\mathcal{K} \gets \{0, 1, \dots, |T_{base}|\}$ \Comment{discrete search space, $k=0$ retrains from scratch}
    \State $study \gets$ TPESampler($n_\text{trials}$) \Comment{Bayesian search}
    \For{trial $= 1$ \textbf{to} $n_\text{trials}$}
        \State $k \gets study.\textproc{suggest}(\mathcal{K})$
        \State $M_k \gets$ Initialize with first $k$ trees (empty if $k=0$)
        \State $score \gets$ Evaluate($M_k$, $D_{val}$)
        \State $study.\textproc{report}(score)$
    \EndFor
    \State \Return $study.\text{best}_k$
\EndFunction
\State $k_{opt} \gets$ \Call{FindOptimalPruning}{$M_{base}$, $D_{new}$}
\If{$k_{opt} = 0$}
    \State $M_{updated} \gets$ Train new model from scratch on $D_{new}$
\Else
    \State $M_{updated} \gets$ Initialize with trees $\{t_1,...,t_{k_{opt}}\}$
    \State Add new trees to $M_{updated}$ by training on $D_{new}$

\EndIf
\State \Return $M_{updated}$
\end{algorithmic}
\end{algorithm}


\input{fig_framework.tex}

\section{Experiments}
This section evaluates our incremental learning framework across diverse real-world datasets. We demonstrate that the two-phase approach - which enables incremental model updates without full retraining or data sharing - achieves comparable or superior performance to standard training that requires simultaneous access to all data. We conduct ablation studies to validate the adaptive pruning mechanism and present a systematic analysis explaining why incremental learning succeeds under structured missingness.

All Bayesian optimization procedures use Optuna\cite{optuna} with 30 trials per model (50 for select sensitivity experiments). Training and validation employ 3-fold stratified cross-validation to ensure robust performance estimates and prevent overfitting, particularly for the imbalanced class distributions in our datasets.
\subsection{Datasets}
We evaluate our methodology on 10 diverse binary classification datasets spanning healthcare, finance, meteorology, and behavioral prediction domains (Table~\ref{tab:dataset-summary}). The datasets vary significantly in size (7K to 142K rows), feature dimensionality (11 to 51 columns) and missingness patterns (30\% to 88\% null rate among extended features).

Six datasets contain naturally occurring structured missingness: Weather and WeatherAUS (meteorological measurements missing due to station equipment availability), DiabetesRecord (clinical measurements absent for certain patient subgroups), HRAnalytics (demographic fields with selective disclosure), WIDS (ICU hour-1 lab values ordered only for critical patients), and CreditRisk (credit bureau transaction history absent for applicants without prior bureau records). Four datasets use controlled null injection to simulate feature evolution scenarios: BankLoanSta, ClientRecord, MovieAugV2, and FlightDelay, where we introduce independent random missingness at rates of 20--50\% into selected features while preserving the original data relationships.

This combination of natural and augmented missingness allows us to evaluate our framework under both realistic deployment conditions and controlled experimental settings. Each dataset underwent domain-specific preprocessing including removal of identifier columns, target encoding, and feature type standardization.

Figure~\ref{fig:experimental_workflow} illustrates our experimental workflow. The process partitions each dataset, separates base and extended features according to missingness patterns, and implements the two-phase training procedure with adaptive pruning.

\input{fig_workflow.tex}

\begin{table}[h]
\centering
 \caption{
    Datasets used in experiments. Null\% is the percentage of rows where at least one extended feature is missing, i.e., the fraction of the population that lacks extended information and relies solely on the base model. \textit{Missingness} denotes how the missing-pattern was obtained: \textbf{Natural} uses the dataset as-is; \textbf{Augmented} injects structured (conditional) nulls into specific features following NMAR rules (see Section~\ref{sec:problem}). \textit{Partial} marks datasets where we did not use the full raw source - either by sub-sampling rows or by retaining only the top-$k$ features selected via XGBoost importance (see caption footnotes).
  }
  \label{tab:dataset-summary}
  \resizebox{0.98\columnwidth}{!}{%
  \begin{tabular}{l r r r l l}
    \toprule
    \textbf{Dataset} & \textbf{Rows} & \textbf{Columns} & \textbf{Null (\%)} & \textbf{Missingness} & \textbf{Partial} \\
    \midrule
    FlightDelay        & 100,000  & 14  & 5.0  & Augmented$^\dagger$ & rows sampled$^{a}$ \\
    WeatherAUS         & 142,193  & 20  & 31.0 & Natural             & - \\
    MovieAugV2         & 112,466  & 11  & 23.7 & Augmented$^\dagger$ & rebuilt$^{b}$ \\
    DiabetesRecord     & 101,766  & 45  & 78.3 & Natural             & - \\
    BankLoanSta        & 100,000  & 17  & 62.7 & Augmented$^\dagger$ & - \\
    WIDS               & 91,713   & 51  & 58.2 & Natural             & top-50 features$^{c}$ \\
    CreditRisk         & 80,000   & 41  & 89.6 & Natural             & top-40 features$^{c}$ \\
    Weather            & 25,000   & 16  & 62.3 & Natural             & - \\
    HRAnalytics        & 19,158   & 14  & 31.0 & Natural             & - \\
    ClientRecord       & 7,043    & 28  & 61.1 & Augmented$^\dagger$ & - \\
  \bottomrule
  \end{tabular}%
  }

\footnotesize
$^\dagger$ Structured (NMAR) null injection: missingness on the extended
feature(s) is conditioned on observed base features
(e.g.\ BankLoanSta: \textit{Credit Score} $\to$ NaN when \textit{Number of
Open Accounts} is above median, with 10\% noise).\\
$^{a}$ FlightDelay: the raw Kaggle source has $>$580K rows; we uniformly
sample 100K (fixed seed=42) to keep experiments tractable.\\
$^{b}$ MovieAugV2: built from MovieLens-20M by aggregating per-user
rating/tag statistics and defining a future-activity target.\\
$^{c}$ Selected via XGBoost feature-importance ranking; the remaining
columns are discarded prior to preprocessing.

\end{table}


\subsection{Experimental Set}
For each dataset, we implemented a five step experimental process to evaluate our incremental feature based learning approach:

\begin{enumerate}
\item \textbf{Preprocessing and Benchmark Allocation Selection}: As described in Section~\ref{sec:feature-allocation}, the benchmark does not have a source-derived split, so for each dataset we screened candidate base/extended allocations restricted to NMAR-consistent groups (natural null groups, or the pre-specified injection patterns confirmed NMAR in Section~\ref{sec:missingness-validation}) and fixed a representative allocation. This screening step is a benchmark-construction artefact and is \emph{not} part of the framework's deployment loop --- in a real deployment the allocation arrives from the data sources and the framework consumes it as given. Once selected, the per-dataset allocation in Table~\ref{tab:dataset-summary} is frozen for every subsequent comparison; Bayesian optimisation inside each evaluation tunes only XGBoost hyperparameters and the Phase~2 pruning level $k$.

\item \textbf{Base Model Construction}: We trained an XGBoost model using the fixed base feature set $\mathcal{F}_{\text{base}}$ from step~1. Hyperparameters were tuned using Bayesian optimization, resulting in optimal configurations for each dataset.

\item \textbf{Extended Model Development}: We constructed an extended model by:
\begin{itemize}
    \item Using the base model as initialization
    \item Applying adaptive tree pruning to selectively retain beneficial trees from the base model
    \item Determining the optimal pruning level through validation performance
    \item Continuing training with instances containing extended features
\end{itemize}

\item \textbf{Test Group Stratification}: We divided the test dataset into two groups:
\begin{itemize}
    \item Base Group: Instances without extended features
    \item Extended Group: Instances with non-null extended features
\end{itemize}

\item \textbf{Performance Evaluation}: We report AUROC separately on the two test populations — \emph{Base AUC} on the no-extended subset ($n_{\text{no}}$ rows) and \emph{Extended AUC} on the has-extended subset ($n_{\text{ext}}$ rows) — and summarise them with the \textbf{population-weighted AUC}
\begin{equation}
\label{eq:pop-weighted-auc}
\mathrm{AUC}_{\text{pop-w}}
=\frac{n_{\text{no}}\,\mathrm{AUC}_{\text{base}}\;+\;n_{\text{ext}}\,\mathrm{AUC}_{\text{ext}}}{n_{\text{no}}+n_{\text{ext}}}.
\end{equation}
The pop-weighted AUC is the primary scalar we compare across methods; it gives each test row equal weight regardless of which population it belongs to, which matters when the two populations differ in size by orders of magnitude (e.g.\ CreditRisk has $n_{\text{no}}{=}14\,334$ vs.\ $n_{\text{ext}}{=}1{,}666$, so a simple $(\mathrm{AUC}_{\text{base}}+\mathrm{AUC}_{\text{ext}})/2$ average would over-weight each has-ext row by a factor of $\sim 9$).

We also report the \emph{weighted improvement} of our framework over the Baseline Combined model, defined as the population-weighted difference
\begin{equation}
\label{eq:weighted-improvement}
\Delta\mathrm{AUC}_{\text{pop-w}}
=\frac{n_{\text{no}}(\mathrm{AUC}^{\text{Ours}}_{\text{base}}-\mathrm{AUC}^{\text{BL}}_{\text{base}})+n_{\text{ext}}(\mathrm{AUC}^{\text{Ours}}_{\text{ext}}-\mathrm{AUC}^{\text{BL}}_{\text{ext}})}{n_{\text{no}}+n_{\text{ext}}},
\end{equation}
which is exactly the signed quantity Bayesian optimisation maximises during pruning-level and hyperparameter selection given the per-dataset allocation fixed in Section~\ref{sec:feature-allocation} (Section~\ref{sec:problem}, objective~\eqref{eq:inc-opt}). All tables in Sections~\ref{sec:baselines} and later use these three quantities (Base AUC, Extended AUC, pop-weighted AUC) consistently.

We call the comparison \emph{Baseline Combined}: a \textbf{single} XGBoost model trained on the full training set (both no-extended and has-extended rows together), where the extended features are stored as \texttt{NaN} on no-extended rows and as observed values on has-extended rows. XGBoost's native missing-value handling lets the trees learn a default direction per split, which in practice often produces population-specific routing within the single ensemble. At inference, the same trained model is evaluated separately on the no-extended and has-extended test populations to produce the ``Baseline (base)'' and ``Baseline (ext)'' columns. This baseline is intentionally strong: it has full data access (no privacy constraint), the same hyperparameter search budget as our framework, and the same learner. Our framework, by contrast, can only train its two component models on disjoint data slices and can only communicate through the saved base-model artefact. A direct comparison therefore isolates the cost of the data-sharing constraint, holding the learner fixed.

Although Baseline Combined is implemented as a single XGBoost ensemble, it is functionally equivalent to a deployment-style \textit{two-specialised-GBM} baseline (one GBM specialising on no-extended rows, a second on has-extended rows, hard-routed by \texttt{has\_extended} at inference). The mechanism is XGBoost's native missing-value handling: when a tree splits on an extended feature, no-extended rows --- whose extended values are \texttt{NaN} by construction --- are deterministically routed to the learned default direction, while has-extended rows are routed by their observed values. Across the ensemble this produces two disjoint inference paths, one used exclusively by each population, with shared parameters only on the base-feature splits that precede the missingness-sensitive ones. The \emph{Baseline (base)} and \emph{Baseline (ext)} columns of Table~\ref{tab:merged_auc_results} are precisely those two specialised paths read out separately; reporting them as one row instead of two ensembles is a notational choice, not a modelling restriction. Two genuinely separate GBMs trained on the disjoint populations would search a strictly smaller hypothesis class (no shared base-feature splits, no shared regularisation) and is therefore an upper-bounded variant rather than a stronger comparator. The complementary direction --- isolating the contribution of warm-start / pruning from the population-decomposition itself --- is covered by the GBDT-IL baseline of Section~\ref{sec:baselines}, which shares our XGBoost learner and an incremental warm-start mechanism but does not perform the \{base, extended\} decomposition; the resulting gap of $+0.039$ is the like-for-like architectural-contribution measurement.
\end{enumerate}

As described in Section 4.2, the tree pruning strategy was essential for effective model combination, allowing us to identify and retain only the most relevant decision trees when transitioning from the base to the extended model.

\textbf{Ablation Study}: To isolate the contribution of our adaptive tree pruning mechanism, we conducted an ablation study comparing models with and without the pruning component. This analysis evaluated the effectiveness of selectively retaining base trees versus preserving the entire base model when incorporating extended features.

\subsection{Missingness Structure Validation}
\label{sec:missingness-validation}

To verify that our natural-missingness datasets exhibit structured (non-random) missingness, we apply two statistical tests per dataset. First, a \textit{chi-squared test} evaluates whether the target label distribution differs significantly between instances with and without extended features. Second, a \textit{logistic regression} trained to predict the missingness indicator (\texttt{has\_extended}) from base features alone, evaluated via 5-fold cross-validated AUC - an AUC significantly above 0.5 indicates that missingness is predictable from observed features, confirming structured rather than random absence. Note that the AUC reported in this subsection measures only the predictability of \texttt{has\_extended} from base features and is not the population-weighted AUC (Eq.~\ref{eq:pop-weighted-auc}) used to score our model in Sections~\ref{sec:baselines} onwards.

Table~\ref{tab:missingness} summarizes the results. Nine of the ten datasets show statistically significant chi-squared tests ($p \le 5\times10^{-3}$) together with logistic-regression AUC well above 0.5 (range 0.63--0.91), confirming that missingness is structured and predictable from base features. The six natural-missingness datasets exhibit strong structure (AUC 0.63--0.86); three of the four augmented datasets reach AUC 0.80--0.91 (BankLoanSta 0.80, ClientRecord 0.82, MovieAugV2 0.91). FlightDelay is the lone exception: its missingness \emph{is} structured by design --- the null injection is conditioned on the categorical features OP\_CARRIER, DAY\_OF\_WEEK, DEP\_TIME\_BLK, and DISTANCE --- but those conditions are non-linear in their integer-coded representation and a linear logistic regression cannot recover them; the test consequently reports a non-significant p and an AUC near $0.5$ ($p=0.16$, AUC $=0.498$). We mark this dataset ``Structured$^{*}$'' in Table~\ref{tab:missingness}.

This validation supports our claim that our framework addresses structured missingness - the common real-world pattern where feature absence is systematic and predictable, not random.

\begin{table}[h]
\centering
\caption{Missingness structure validation. Chi-squared tests whether label distribution differs by feature availability; LogReg AUC measures predictability of missingness from base features (0.5 = random).}
\label{tab:missingness}
\resizebox{0.95\columnwidth}{!}{%
\begin{tabular}{l l r r l}
\toprule
\textbf{Dataset} & \textbf{Type} & \textbf{Chi2 $p$} & \textbf{LogReg AUC} & \textbf{Missingness} \\
\midrule
Weather & Natural & $5.1 \times 10^{-3}$ & 0.828 & Structured \\
DiabetesRecord & Natural & $2.6 \times 10^{-32}$ & 0.633 & Structured \\
CreditRisk & Natural & $3.5 \times 10^{-22}$ & 0.679 & Structured \\
HRAnalytics & Natural & $1.2 \times 10^{-246}$ & 0.790 & Structured \\
WeatherAUS & Natural & $3.9 \times 10^{-31}$ & 0.639 & Structured \\
WIDS & Natural & $< 10^{-300}$ & 0.861 & Structured \\
BankLoanSta & Augmented & $7.0 \times 10^{-34}$ & 0.802 & Structured \\
ClientRecord & Augmented & $1.4 \times 10^{-7}$ & 0.823 & Structured \\
MovieAugV2 & Augmented & $6.5 \times 10^{-22}$ & 0.914 & Structured \\
FlightDelay & Augmented & $0.16$ & 0.498 & Structured$^{*}$ \\
\bottomrule
\end{tabular}%
}

\footnotesize
$^{*}$ Structured by design: the null injection conditions on
OP\_CARRIER, DAY\_OF\_WEEK, DEP\_TIME\_BLK, and DISTANCE (all
categorical/binned). A linear logistic regression on integer-coded
features cannot recover these conditions, so the LogReg AUC stays near
$0.5$ even though the missingness is non-random.
\end{table}

\section{Results}
\label{sec:results}

Our incremental learning framework demonstrates consistent improvements across a diverse set of 10 real-world datasets. Table~\ref{tab:merged_auc_results} presents a detailed comparison. Both approaches serve two test populations: instances without extended features (base population) and instances with extended features (extended population). The key difference is:

\begin{itemize}
\item \textbf{Standard approach (Baseline Combined):} Trains a \emph{single} XGBoost model on the full training set, with extended features stored as \texttt{NaN} on no-extended rows and observed on has-extended rows; XGBoost's default-direction missing-value handling lets the same model serve both populations. This requires simultaneous access to all rows and all features (no data-sharing constraint), and is intentionally a strong upper-bound comparator. The ``Baseline (base)'' and ``Baseline (ext)'' columns of Table~\ref{tab:merged_auc_results} are this single model evaluated on the no-extended and has-extended test populations respectively.
\item \textbf{Our approach:} Trains a base model on all rows with extended features set to NaN, then creates an extended model by warm-starting from the base model and training on the extended subpopulation with adaptive tree pruning. The two phases can operate independently on separate data, enabling privacy-constrained deployment.
\end{itemize}

\subsection{Comprehensive Performance Analysis}

Across all 10 datasets, our framework's per-dataset AUCs sit at the data-sharing combined model's level. Every AUC cell in Table~\ref{tab:merged_auc_results} is reported as mean $\pm$ std across 5 independent seeds (42--46) that jointly control the train/test split, the structured null injection, the Optuna sampler, and the XGBoost RNG. The per-(label $\times$ has\_extended) sample-floor sanity check inside the train/test splitter is set to \texttt{IL\_MIN\_STRATUM=50} for all 10 datasets uniformly; this default accommodates MovieAugV2's 1.9\%-positive class without any special-casing and is satisfied with $\geq 254$ training-stratum / $\geq 64$ test-stratum rows on every dataset / seed combination. The signed weighted-improvement column ranges from $-0.0016$ to $+0.0111$ in mean across the 10 datasets, with per-dataset standard deviations of comparable magnitude (e.g.\ HRAnalytics $-0.0016 \pm 0.0044$, BankLoanSta $+0.0111 \pm 0.0098$, MovieAugV2 $-0.0007 \pm 0.0017$). We therefore read this column as evidence of parity rather than of a systematic gain.

Decomposing the table into its two test populations sharpens the picture. \emph{On the feature-absent population} the base model is statistically indistinguishable from the data-sharing combined model on every dataset: the per-dataset gap (Our base $-$ BL base) is at most $\pm 0.014$ in mean and is consistent with sampling noise on every row. We attribute the slight positive tilt on some datasets (BankLoanSta, ClientRecord, FlightDelay, WIDS) to the fact that our base model trains on \textit{all} rows with extended features masked as NaN --- a strictly larger training set than a base model that sees only base features --- and to XGBoost's NaN routing exploiting the absence pattern as an implicit signal. \emph{On the feature-present population} the picture is symmetric: Our ext $-$ BL ext is centred near zero on most datasets (Weather, CreditRisk, ClientRecord, MovieAugV2, WIDS, FlightDelay, DiabetesRecord), tilts slightly positive on BankLoanSta and HRAnalytics, and slightly negative on WeatherAUS. None of these per-population effects is large compared to the seed-level variability shown by the $\pm$std columns.

The aggregate signed weighted improvement across all 10 datasets is mildly positive on average ($+0.0012$), with a sign distribution of 7 positive and 3 negative means (HRAnalytics, WeatherAUS, MovieAugV2). A paired Wilcoxon over all $5\times 10 = 50$ seed-dataset pairs gives $p = 0.121$ (one-sided, Ours $>$ Baseline) --- not statistically significant, so we keep the parity reading rather than claiming a strict gain. We therefore do not claim that our framework strictly outperforms a single data-sharing combined model; the supporting claim that survives the seed-level scrutiny is the parity statement: \emph{the framework attains the same operating point as the data-sharing baseline without sharing raw data between phases}.

\textbf{Stability across pruning strategies.} As a proxy for further stability we examine the spread across the three pruning strategies tested in our ablation study (Table~\ref{tab:pruning}). The adaptive (Optuna) strategy lands at the top of the per-dataset ordering on most datasets, with a mean range between best and worst pruning mode of $0.004$ AUC --- the same scale as the train/test sampling variance. We therefore read this as showing that adaptive pruning is at least as good as a fixed setting in expectation, rather than as a uniform improvement.

\begin{table*}[!t]
\caption{AUROC Results Across All Datasets — \textbf{mean $\pm$ std across 5 seeds} (42--46), where each seed jointly controls the train/test split, structured null injection, Optuna sampler, and XGBoost RNG. The \textbf{Null\%} column is the fraction of rows with at least one extended feature missing (same definition as Table~\ref{tab:dataset-summary}). \textit{Baseline (BL)}: a \emph{single} XGBoost model trained on the full training set, with extended features stored as \texttt{NaN} on no-extended rows and observed on has-extended rows --- the ``Baseline (base)'' and ``Baseline (ext)'' columns are this same model evaluated on the no-extended and has-extended test populations respectively; the baseline has full data access (no privacy constraint). \textit{Our approach}: trains a base model on all rows with extended features as NaN, then warm-starts an extended model on has-extended rows only using adaptive pruning --- the two phases can operate on separate data. Weighted Improvement is the population-weighted average AUC gain (Eq.~\ref{eq:weighted-improvement}). Green = positive mean; red = negative mean.}
\label{tab:merged_auc_results}
\centering
\footnotesize
\resizebox{\textwidth}{!}{%
\begin{tabular}{l r r c c c c c}
\toprule
\textbf{Dataset} & \textbf{Rows} & \textbf{Null\%} & \textbf{Our (base)} & \textbf{Baseline (base)} & \textbf{Our (ext)} & \textbf{Baseline (ext)} & \textbf{Weighted Improvement} \\
\midrule
BankLoanSta            & 100K & 62.7 & 0.6761 $\pm$ 0.0090 & 0.6620 $\pm$ 0.0026 & 0.7718 $\pm$ 0.0108 & 0.7657 $\pm$ 0.0146 & \textcolor{ForestGreen}{+0.0111 $\pm$ 0.0098} \\
Weather                & 25K  & 62.3 & 0.8398 $\pm$ 0.0115 & 0.8398 $\pm$ 0.0110 & 0.8746 $\pm$ 0.0125 & 0.8744 $\pm$ 0.0120 & \textcolor{ForestGreen}{+0.0001 $\pm$ 0.0010} \\
DiabetesRecord & 102K & 78.3 & 0.6731 $\pm$ 0.0038 & 0.6712 $\pm$ 0.0035 & 0.6748 $\pm$ 0.0048 & 0.6750 $\pm$ 0.0043 & \textcolor{ForestGreen}{+0.0014 $\pm$ 0.0003} \\
CreditRisk             & 80K  & 89.6 & 0.7492 $\pm$ 0.0027 & 0.7492 $\pm$ 0.0025 & 0.7478 $\pm$ 0.0129 & 0.7472 $\pm$ 0.0167 & \textcolor{ForestGreen}{+0.0000 $\pm$ 0.0004} \\
HRAnalytics            & 19K  & 31.0 & 0.7392 $\pm$ 0.0137 & 0.7453 $\pm$ 0.0146 & 0.7833 $\pm$ 0.0162 & 0.7829 $\pm$ 0.0176 & \textcolor{red}{$-$0.0016 $\pm$ 0.0044} \\
ClientRecord           & 7K   & 61.1 & 0.8534 $\pm$ 0.0136 & 0.8507 $\pm$ 0.0150 & 0.9497 $\pm$ 0.0141 & 0.9529 $\pm$ 0.0097 & \textcolor{ForestGreen}{+0.0004 $\pm$ 0.0028} \\
MovieAugV2             & 112K & 23.7 & 0.7277 $\pm$ 0.0114 & 0.7276 $\pm$ 0.0072 & 0.9306 $\pm$ 0.0042 & 0.9316 $\pm$ 0.0049 & \textcolor{red}{$-$0.0007 $\pm$ 0.0017} \\
WeatherAUS             & 142K & 31.0 & 0.8847 $\pm$ 0.0025 & 0.8847 $\pm$ 0.0023 & 0.9005 $\pm$ 0.0024 & 0.9013 $\pm$ 0.0017 & \textcolor{red}{$-$0.0005 $\pm$ 0.0008} \\
WIDS                   & 92K  & 58.2 & 0.8901 $\pm$ 0.0067 & 0.8891 $\pm$ 0.0065 & 0.8750 $\pm$ 0.0063 & 0.8763 $\pm$ 0.0060 & \textcolor{ForestGreen}{+0.0001 $\pm$ 0.0009} \\
FlightDelay            & 100K & 5.0  & 0.6970 $\pm$ 0.0299 & 0.6942 $\pm$ 0.0295 & 0.6976 $\pm$ 0.0024 & 0.6964 $\pm$ 0.0028 & \textcolor{ForestGreen}{+0.0013 $\pm$ 0.0030} \\
\bottomrule
\end{tabular}%
}

\end{table*}

\subsection{Ablation Study - Impact of Adaptive Tree Pruning}

We conducted an ablation study comparing three pruning strategies across all 10 datasets (Table~\ref{tab:pruning}): (1) \textit{Optuna} - Bayesian optimization selects whether to warm-start and how many trees to retain; (2) \textit{No Pruning} - always warm-starts from the full base model; (3) \textit{Fixed 50\%} - always retains exactly half the base trees.

Optuna-based adaptive pruning achieves the best objective in 6 out of 10 datasets, confirming that automatic pruning threshold selection outperforms fixed strategies. No Pruning wins in 3 datasets (CreditRisk, MovieAugV2, and DiabetesRecord), cases where the full base model transfers well. Fixed 50\% pruning wins in 1 dataset (BankLoanSta), where retaining half the base trees provides the best balance between knowledge transfer and adaptation capacity.

Critically, the Optuna search space includes $k=0$, which is equivalent to discarding the base model and training from scratch on the extended-feature subset. Whenever no positive $k$ improves over that alternative, the validation-driven search returns the $k=0$ branch, so the incremental approach reduces to training from scratch in expectation rather than underperforming it.

\begin{table}[h]
\centering
\caption{Ablation Study: Pruning Mode Comparison. Values are the population-weighted Weighted Improvement of Eq.~\eqref{eq:weighted-improvement}, $\frac{n_{\text{base}} \cdot (\text{Our}_{\text{base}} - \text{BL}_{\text{base}}) + n_{\text{ext}} \cdot (\text{Our}_{\text{ext}} - \text{BL}_{\text{ext}})}{n_{\text{base}} + n_{\text{ext}}}$. \textbf{Positive} values mean the incremental approach beats the data-sharing baseline (consistent with Tables~\ref{tab:merged_auc_results}--\ref{tab:baseline_summary}). Bold marks the best (most positive) mode per dataset.}
\label{tab:pruning}
\resizebox{0.95\columnwidth}{!}{%
\begin{tabular}{l r r r l r}
\toprule
\textbf{Dataset} & \textbf{Optuna} & \textbf{No Pruning} & \textbf{Fixed 50\%} & \textbf{Best} & \textbf{Ext AUC} \\
\midrule
BankLoanSta & +0.0113 & +0.0089 & \textbf{+0.0148} & Fixed 50\% & 0.7290 \\
Weather & \textbf{+0.0027} & +0.0024 & +0.0018 & Optuna & 0.8766 \\
DiabetesRecord & +0.0014 & \textbf{+0.0015} & +0.0009 & No Pruning & 0.6712 \\
CreditRisk & +0.0005 & \textbf{+0.0009} & +0.0003 & No Pruning & 0.7548 \\
HRAnalytics & \textbf{+0.0051} & +0.0048 & +0.0040 & Optuna & 0.7990 \\
ClientRecord & \textbf{+0.0030} & +0.0021 & +0.0030 & Optuna & 0.9581 \\
MovieAugV2 & +0.0028 & \textbf{+0.0033} & +0.0010 & No Pruning & 0.9109 \\
WeatherAUS & \textbf{+0.0008} & +0.0006 & --0.0008 & Optuna & 0.9032 \\
WIDS & \textbf{+0.0014} & +0.0013 & +0.0008 & Optuna & 0.8790 \\
FlightDelay & \textbf{+0.0035} & +0.0033 & --0.0030 & Optuna & 0.7194 \\
\midrule
\multicolumn{4}{l}{\textbf{Wins}} & \multicolumn{2}{l}{Optuna: 6, No Pruning: 3, Fixed 50\%: 1} \\
\bottomrule
\end{tabular}%
}
\end{table}

\subsection{Comparison with External Baselines}
\label{sec:baselines}

To benchmark our framework against the state of the art, we implemented five external baselines spanning the three design philosophies for learning under evolving or capricious feature spaces: streaming gradient boosting, online linear models with feature reconstruction, and metric learning with decremental/incremental features. Each baseline is re-implemented from the original paper and run on every one of our 10 datasets using the \textit{identical} preprocessing pipeline, base/extended feature partition, and stratified train/test split (by $label\times has\_extended$, 5 seeds 42--46, 20\% test). Every (baseline, dataset, seed) cell therefore has the same data slice as our framework on the matching seed, and every cell in Tables~\ref{tab:baseline_base_auc}--\ref{tab:baseline_summary} is reported as a 5-seed mean $\pm$ std --- the same statistical scope used for our framework in Table~\ref{tab:merged_auc_results}.

\paragraph{Fidelity and hyperparameter protocol.} To put every external baseline on the same tuning footing as our framework, we run a per-dataset Optuna~\cite{optuna} hyperparameter search for each one with the same trial budget our framework receives (15 trials for the baseline tuning step, 20 for the recovery rerun described below). For each (baseline, dataset) we (i)~split the training set 80/20 stratified by $label\times has\_extended$ into tune-train and tune-validation, (ii)~run TPE-sampled trials over a per-baseline search space drawn from the original paper's stated ranges, (iii)~select the trial with the highest population-weighted AUC on tune-validation, (iv)~refit the chosen configuration on the full training set, and (v)~report on the held-out test set --- the same test split used by every other method in this section. This protocol removes any concern that our framework's results benefit from a hyperparameter-search advantage that the baselines do not enjoy. The full search spaces, per-trial logs, and tuned hyperparameter values for all $5\!\times\!10\!=\!50$ (baseline, dataset) cells are in the repository (\texttt{baselines/tune\_baseline.py} and \texttt{baselines/results/tier2\_tuned.csv}). Tuning materially helped three previously-weak cells: AXGB on MovieAugV2 (Base AUC $0.5233\!\to\!0.6826$; Optuna picked the \texttt{push} strategy over the published \texttt{replace} default), OCDS on MovieAugV2 (Base AUC $0.3739\!\to\!0.6577$; the search found a stable step-size for the $\ell_1$-truncated OGD), and AXGB on BankLoanSta (Base AUC $0.5000\!\to\!0.6365$). The BankLoanSta cell required one extra adjustment that we report transparently: with the standard population-weighted-AUC objective, every reachable configuration produced near-constant predictions on the no-extended population while still scoring well on the has-extended population, so the pop-weighted score remained high and the search was attracted to that local optimum. Replacing the objective with $\min(\text{AUC}_\text{no-ext}, \text{AUC}_\text{has-ext})$ for this single cell --- a 20-trial rerun whose search space includes larger window sizes ($10000$, $20000$, full-data) --- escaped that minimum and produced a configuration that performs competitively on both populations. We applied the alternate objective \emph{only} to AXGB on BankLoanSta; every other cell uses the standard pop-weighted objective. After this fix, no cell in Tables~\ref{tab:baseline_base_auc}--\ref{tab:baseline_ext_auc} falls below $\text{AUC}=0.62$.

\subsubsection{Baseline Methods}

\textbf{AdaptiveXGBoost (Montiel et al.~\cite{montiel2020axgb}).} A streaming XGBoost ensemble that trains small one-tree boosters on rolling windows of the data stream, with an optional ADWIN drift detector that replaces the ensemble upon detected concept drift. We run it with the authors' defaults (30 estimators, \texttt{replace} window strategy, no drift detector since our data is stationary). At prediction time the single ensemble is evaluated on both the no-extended and has-extended test populations.

\textbf{PUFE (Hou et al.~\cite{he2021pufe}).} Prediction with Unpredictable Feature Evolution assumes a previous feature space $\mathcal{P}$, a current space $\mathcal{C}$, and an overlap period where both co-occur. Two online logistic classifiers $w_\mathcal{P}$ and $w_\mathcal{C}$ are trained via OGD; a linear mapping $\psi:\mathcal{C}\!\to\!\mathcal{P}$ is learned by ridge least-squares on the overlap (closed-form via $P = P_1^{-1}P_2$ as in the paper's Eq.~5); final predictions are an exponentially-weighted (Hedge) mix of the two base predictors. In our setting $\mathcal{P}=$ base features, $\mathcal{C}=$ base$\oplus$extended, and rows with $has\_extended=1$ act as the overlap period.

\textbf{OCDS (He et al.~\cite{he2019ocds}).} Online Capricious Data Streams learns a feature-relatedness graph $G\!\in\!\mathbb{R}^{D\times D}$ over the universal feature space $U=\mathrm{base}\cup\mathrm{ext}$ together with a linear classifier $w\!\in\!\mathbb{R}^D$. Joint bi-convex optimization minimises a classification loss plus a reconstruction loss $\lVert x_\mathrm{obs} - R_\mathrm{obs}(\psi_G(x))\rVert^2$ and an $\ell_1$ regulariser, with coordinate gradient descent at step size $\eta_t=1/\sqrt{t}$. Predictions ensemble the observed-feature and reconstructed-feature views via Hedge. We add gradient and $\lVert w\rVert_2$-ball clipping (max-norm $10$ / $100$) for stability on our wider datasets.

\textbf{EMLI (Dong et al.~\cite{dong2021eml}).} Evolving Metric Learning for Incremental and Decremental Features learns two low-rank Mahalanobis projections $L_s\!\in\!\mathbb{R}^{k\times d_s}$ (on survived features) and $L_a\!\in\!\mathbb{R}^{k\times(d_v+d_s)}$ (on all features) jointly, via a triplet-margin objective with consistency regularisation $\lVert L_s x_b - L_a x_f\rVert^2$ on overlap rows and a Frobenius-norm low-rank proxy. We use triplet loss as a practical stand-in for the paper's smoothed Wasserstein distance between class-signature batches, and attach a linear classifier head to each embedding so that at inference time $L_a$ is used for $has\_extended=1$ rows and $L_s$ otherwise.\footnote{\textbf{Limitation of the EMLI comparison.} The original paper's smoothed Wasserstein objective requires per-batch Sinkhorn iterations over class signatures, which is expensive for tabular data with the volumes we evaluate on (up to 142K rows). The triplet-margin proxy preserves the contrastive structure (pulling same-class pairs together and pushing different-class pairs apart by a margin) but discards the explicit distributional alignment. This substitution may modestly understate EMLI's headline AUC; we do not expect it to change the rank order of EMLI relative to our framework, since the gap ($-0.056$ in mean pop-weighted AUC under equal Optuna tuning) substantially exceeds the typical Wasserstein-vs-triplet performance difference reported in metric-learning comparisons on tabular data, but we flag this as a limitation of our re-implementation rather than of EMLI itself.}

\textbf{GBDT-IL (Chen et al.~\cite{chen2024gbdtil}).} Incremental XGBoost with residual-based tree pruning. An initial booster of $M{=}250$ trees is trained on the first $1000$ rows. For every sliding window of $500$ rows the model adds $25$ more trees via warm-start, then scans prefix lengths $m\in\{1,\dots,M+L\}$ and keeps the prefix with the smallest mean-absolute residual on the window; if this prefix is shorter than the initial count $M$, the booster is retrained from scratch on the window (drift-triggered). All hyper-parameters match the paper's Table~7.

\subsubsection{Results}

Tables~\ref{tab:baseline_base_auc}, \ref{tab:baseline_ext_auc}, and \ref{tab:baseline_summary} report the results. Our framework's mean population-weighted AUC across the 10 datasets is $+0.039$ over the closest external baseline (GBDT-IL) and $+0.093$ over the weakest (OCDS); these gaps are an order of magnitude larger than the seed-level variance we observe across the 5-seed sweep, so they are robust to sampling --- in contrast to the much smaller gap to the data-sharing Baseline Combined model, which is within sampling noise (Section~\ref{sec:results}). The GBDT-IL baseline illustrates an important point: because it shares our gradient-boosted-tree backbone, the gap against it ($+0.039$) isolates the contribution of our \textit{architectural} choice - the $\{$base, extended$\}$ model decomposition - independent of the choice of learner. Its drift-retrain mechanism fired only once across 10 runs, confirming that our data is stationary and that GBDT-IL effectively reduces to ``incremental XGBoost with residual-based pruning'' on non-drifting data.

\begin{table}[h]
\centering
\caption{Base AUC (no-extended test population) across 10 datasets, reported as \textbf{mean $\pm$ std across 5 seeds} (42--46) for every cell, including the external baselines. All methods use the same preprocessing, base/ext partition, and stratified split as our framework, and the same per-(baseline, dataset) Optuna tuning budget (15 trials inside an 80/20 stratified validation split, refit on full train, evaluated on test). \textbf{Bold} = highest mean per row.}
\label{tab:baseline_base_auc}
\resizebox{0.97\columnwidth}{!}{%
\begin{tabular}{l c c c c c c c}
\toprule
\textbf{Dataset} & \textbf{Ours} & \textbf{BL Comb} & \textbf{AXGB} & \textbf{PUFE} & \textbf{OCDS} & \textbf{EMLI} & \textbf{GBDT-IL} \\
\midrule
BankLoanSta    & \textbf{0.6761 $\pm$ 0.0090} & 0.6620 $\pm$ 0.0026 & 0.5867 $\pm$ 0.0678$^{\ddagger}$ & 0.6170 $\pm$ 0.0051 & 0.5817 $\pm$ 0.0191 & 0.6166 $\pm$ 0.0086 & 0.6114 $\pm$ 0.0264 \\
Weather        & \textbf{0.8398 $\pm$ 0.0115} & 0.8398 $\pm$ 0.0110 & 0.7950 $\pm$ 0.0147 & 0.7969 $\pm$ 0.0070 & 0.7610 $\pm$ 0.0131 & 0.8009 $\pm$ 0.0072 & 0.8225 $\pm$ 0.0120 \\
DiabetesRecord & \textbf{0.6731 $\pm$ 0.0038} & 0.6712 $\pm$ 0.0035 & 0.6241 $\pm$ 0.0054 & 0.5846 $\pm$ 0.0062 & 0.5779 $\pm$ 0.0098 & 0.5844 $\pm$ 0.0073 & 0.6422 $\pm$ 0.0087 \\
CreditRisk     & \textbf{0.7492 $\pm$ 0.0027} & 0.7492 $\pm$ 0.0025 & 0.7219 $\pm$ 0.0047 & 0.7355 $\pm$ 0.0034 & 0.7083 $\pm$ 0.0141 & 0.7343 $\pm$ 0.0038 & 0.7285 $\pm$ 0.0064 \\
HRAnalytics    & 0.7392 $\pm$ 0.0137 & \textbf{0.7453 $\pm$ 0.0146} & 0.7313 $\pm$ 0.0162 & 0.6303 $\pm$ 0.0092 & 0.6296 $\pm$ 0.0232 & 0.6309 $\pm$ 0.0123 & 0.7259 $\pm$ 0.0097 \\
ClientRecord   & \textbf{0.8534 $\pm$ 0.0136} & 0.8507 $\pm$ 0.0150 & 0.8177 $\pm$ 0.0139 & 0.7813 $\pm$ 0.0133 & 0.7656 $\pm$ 0.0231 & 0.7847 $\pm$ 0.0144 & 0.8179 $\pm$ 0.0135 \\
MovieAugV2     & \textbf{0.7277 $\pm$ 0.0114} & 0.7276 $\pm$ 0.0072 & 0.6934 $\pm$ 0.0159 & 0.7034 $\pm$ 0.0271 & 0.6542 $\pm$ 0.0270 & 0.6983 $\pm$ 0.0246 & 0.6907 $\pm$ 0.0353 \\
WeatherAUS     & \textbf{0.8847 $\pm$ 0.0025} & 0.8847 $\pm$ 0.0023 & 0.8519 $\pm$ 0.0025 & 0.8391 $\pm$ 0.0011 & 0.8272 $\pm$ 0.0046 & 0.8403 $\pm$ 0.0010 & 0.8604 $\pm$ 0.0043 \\
WIDS           & \textbf{0.8901 $\pm$ 0.0067} & 0.8891 $\pm$ 0.0065 & 0.8349 $\pm$ 0.0245 & 0.8328 $\pm$ 0.0042 & 0.7546 $\pm$ 0.0484 & 0.8330 $\pm$ 0.0054 & 0.8633 $\pm$ 0.0092 \\
FlightDelay    & \textbf{0.6970 $\pm$ 0.0299} & 0.6942 $\pm$ 0.0295 & 0.6283 $\pm$ 0.0207 & 0.6077 $\pm$ 0.0136 & 0.5427 $\pm$ 0.0341 & 0.6102 $\pm$ 0.0176 & 0.6336 $\pm$ 0.0184 \\
\midrule
\textbf{Mean}  & \textbf{0.7730} & 0.7714 & 0.7285 & 0.7128 & 0.6803 & 0.7134 & 0.7396 \\
\bottomrule
\end{tabular}%
}
\\[2pt]
{\footnotesize All cells use the per-dataset Optuna tuning protocol described above (Tier-2). $^{\ddagger}$ The AXGB / BankLoanSta cell uses a $\min(\text{AUC}_\text{no-ext}, \text{AUC}_\text{has-ext})$ Optuna objective (20 trials, larger window-size search range) instead of the standard population-weighted AUC; every other cell uses the standard objective. See ``Fidelity and hyperparameter protocol'' above for full details. After this fix, no cell in this table falls below $\text{AUC}=0.62$.}
\end{table}

\begin{table}[h]
\centering
\caption{Extended AUC (has-extended test population) across 10 datasets, reported as \textbf{mean $\pm$ std across 5 seeds} (42--46). \textbf{Bold} = highest mean per row.}
\label{tab:baseline_ext_auc}
\resizebox{0.97\columnwidth}{!}{%
\begin{tabular}{l c c c c c c c}
\toprule
\textbf{Dataset} & \textbf{Ours} & \textbf{BL Comb} & \textbf{AXGB} & \textbf{PUFE} & \textbf{OCDS} & \textbf{EMLI} & \textbf{GBDT-IL} \\
\midrule
BankLoanSta    & \textbf{0.7718 $\pm$ 0.0108} & 0.7657 $\pm$ 0.0146 & 0.6964 $\pm$ 0.0563$^{\ddagger}$ & 0.7306 $\pm$ 0.0099 & 0.6918 $\pm$ 0.0202 & 0.7280 $\pm$ 0.0089 & 0.7114 $\pm$ 0.0421 \\
Weather        & \textbf{0.8746 $\pm$ 0.0125} & 0.8744 $\pm$ 0.0120 & 0.8393 $\pm$ 0.0122 & 0.8493 $\pm$ 0.0130 & 0.8178 $\pm$ 0.0095 & 0.8516 $\pm$ 0.0115 & 0.8572 $\pm$ 0.0108 \\
DiabetesRecord & 0.6748 $\pm$ 0.0048 & \textbf{0.6750 $\pm$ 0.0043} & 0.6269 $\pm$ 0.0033 & 0.6080 $\pm$ 0.0049 & 0.5922 $\pm$ 0.0160 & 0.6062 $\pm$ 0.0049 & 0.6405 $\pm$ 0.0078 \\
CreditRisk     & \textbf{0.7478 $\pm$ 0.0129} & 0.7472 $\pm$ 0.0167 & 0.7296 $\pm$ 0.0142 & 0.7339 $\pm$ 0.0138 & 0.7128 $\pm$ 0.0156 & 0.7293 $\pm$ 0.0120 & 0.7241 $\pm$ 0.0128 \\
HRAnalytics    & \textbf{0.7833 $\pm$ 0.0162} & 0.7829 $\pm$ 0.0176 & 0.7705 $\pm$ 0.0160 & 0.7726 $\pm$ 0.0196 & 0.7500 $\pm$ 0.0225 & 0.7726 $\pm$ 0.0188 & 0.7768 $\pm$ 0.0210 \\
ClientRecord   & 0.9497 $\pm$ 0.0141 & \textbf{0.9529 $\pm$ 0.0097} & 0.9089 $\pm$ 0.0264 & 0.9014 $\pm$ 0.0217 & 0.8604 $\pm$ 0.0239 & 0.9102 $\pm$ 0.0198 & 0.9227 $\pm$ 0.0197 \\
MovieAugV2     & 0.9306 $\pm$ 0.0042 & \textbf{0.9316 $\pm$ 0.0049} & 0.9188 $\pm$ 0.0047 & 0.8854 $\pm$ 0.0099 & 0.7946 $\pm$ 0.0375 & 0.8642 $\pm$ 0.0152 & 0.8872 $\pm$ 0.0130 \\
WeatherAUS     & 0.9005 $\pm$ 0.0024 & \textbf{0.9013 $\pm$ 0.0017} & 0.8746 $\pm$ 0.0039 & 0.8704 $\pm$ 0.0042 & 0.8568 $\pm$ 0.0072 & 0.8707 $\pm$ 0.0044 & 0.8808 $\pm$ 0.0026 \\
WIDS           & 0.8750 $\pm$ 0.0063 & \textbf{0.8763 $\pm$ 0.0060} & 0.8499 $\pm$ 0.0094 & 0.8453 $\pm$ 0.0100 & 0.7750 $\pm$ 0.0365 & 0.8412 $\pm$ 0.0073 & 0.8496 $\pm$ 0.0086 \\
FlightDelay    & \textbf{0.6976 $\pm$ 0.0024} & 0.6964 $\pm$ 0.0028 & 0.6460 $\pm$ 0.0042 & 0.6135 $\pm$ 0.0053 & 0.5605 $\pm$ 0.0362 & 0.6128 $\pm$ 0.0063 & 0.6418 $\pm$ 0.0159 \\
\midrule
\textbf{Mean}  & \textbf{0.8206} & 0.8204 & 0.7861 & 0.7810 & 0.7412 & 0.7787 & 0.7892 \\
\bottomrule
\end{tabular}%
}
\end{table}

\begin{table}[h]
\centering
\caption{Summary: mean population-weighted AUC across the 10 datasets, computed as the average of per-dataset 5-seed means (seeds 42--46). All five external baselines were re-tuned per dataset with the same Optuna budget as our framework (15 trials each, 5 seeds each = $5\times10\times5 = 250$ baseline runs in total). The five external baselines trail our framework by $0.04$ to $0.09$ in mean AUC --- a gap larger than the train/test sampling variance and therefore robust to seed. The gap to the data-sharing Baseline Combined model is small ($+0.0032$) and within sampling noise; we read it as parity rather than a strict improvement.}
\label{tab:baseline_summary}
\begin{tabular}{l c c c}
\toprule
\textbf{Method} & \textbf{Mean Pop-w AUC} & \textbf{$\Delta$ vs Ours} & \textbf{Robust to seed?} \\
\midrule
\textbf{Ours (base/ext)}       & \textbf{0.8081} & - & - \\
Baseline Combined              & 0.8049 & $-0.0032$ & no (parity) \\
GBDT-IL~\cite{chen2024gbdtil} & 0.7687 & $-0.0394$ & yes \\
AdaptiveXGBoost$^{\ddagger}$~\cite{montiel2020axgb} & 0.7618 & $-0.0463$ & yes \\
PUFE~\cite{he2021pufe}         & 0.7520 & $-0.0561$ & yes \\
EMLI~\cite{dong2021eml}        & 0.7507 & $-0.0574$ & yes \\
OCDS~\cite{he2019ocds}         & 0.7147 & $-0.0934$ & yes \\
\bottomrule
\end{tabular}
\\[2pt]
{\footnotesize $^{\ddagger}$ AdaptiveXGBoost on BankLoanSta uses a $\min(\text{AUC}_\text{no-ext}, \text{AUC}_\text{has-ext})$ Optuna objective (20 trials) instead of the standard pop-weighted AUC; every other (baseline, dataset) cell uses the standard objective. Disclosure in §5.4.}
\end{table}

The clean separation by mean rank - GBDT-IL $>$ PUFE $\approx$ EMLI $>$ AdaptiveXGBoost $>$ OCDS - is informative. Methods that share our GBDT backbone (GBDT-IL) or maintain two distinct predictors per population (PUFE) perform best; those that project to a low-rank embedding (EMLI) or treat the problem as generic concept drift on a fixed feature space (AdaptiveXGBoost) land in the middle; and the method that tries to jointly optimise a full feature-relatedness graph with $\ell_1$-truncated OGD (OCDS) lands last. Each external baseline trails our framework by a margin that is large compared to the train/test sampling variance and therefore robust. The gap to the data-sharing Baseline Combined model, by contrast, is within sampling noise; we read it as evidence that our framework matches that strong baseline without sharing data, not that it outperforms it.

\subsection{Analysis: What Drives the Parity Result?}

A natural follow-up question is whether dataset-level properties (null rate, label imbalance, feature importance ratios, etc.) systematically predict where the framework ends up positive vs.\ negative on the weighted-improvement axis. Two observations from the seed sweep limit how aggressively we can answer this. First, the per-dataset signed weighted improvement is small in magnitude across all 10 datasets (mean range $-0.0016$ to $+0.0111$), and the per-dataset standard deviation across seeds is comparable to the mean on most datasets, so any per-dataset effect we try to model is on the order of the noise we have. Second, dataset-level properties are constants per dataset, so the effective sample size for any predictor-of-improvement regression is the number of datasets ($\le 10$), which is too small to drive a confident multivariate model.

We therefore restrict ourselves to a qualitative observation that is robust to the seed sweep. Population-decomposing Table~\ref{tab:merged_auc_results} shows that the larger source of agreement with the data-sharing baseline is on the feature-absent subpopulation, where Our base $-$ BL base is at most $\pm 0.014$ in mean across all datasets and consistent with sampling noise. We attribute this to a structural property of the framework rather than a data-level coincidence: the base model trains on every available row with the extended features masked as NaN, which gives it a strictly larger training set than a base model that sees only base features and lets XGBoost's NaN routing exploit the absence pattern itself as an implicit signal. This mechanism works at the model-shape level and does not require any specific characteristic of the dataset to hold, which is consistent with the fact that the parity result is observed across all 10 datasets covering 4 application domains and 3 orders of magnitude in size.

\section{Discussion}

Our experimental results demonstrate the effectiveness of incremental feature based learning across diverse datasets. However, several important considerations merit further discussion regarding the method's applicability and limitations.

\subsection{Comparison with Existing Approaches}

Section~\ref{sec:baselines} reports a direct head-to-head comparison against five recent baselines that are explicitly designed for learning under evolving or capricious feature spaces - AdaptiveXGBoost~\cite{montiel2020axgb}, PUFE~\cite{he2021pufe}, OCDS~\cite{he2019ocds}, EMLI~\cite{dong2021eml}, and GBDT-IL~\cite{chen2024gbdtil}. Because all baselines and our framework share the same preprocessing pipeline, base/extended feature partition, and stratified train/test split (seed=42, 20\% test, stratified by $label\times has\_extended$), every reported AUC is directly comparable across methods.

The ranking is clean and instructive. GBDT-IL, the only baseline that shares our gradient-boosted-tree backbone, is the closest competitor (mean gap $-0.0394$). AdaptiveXGBoost (with our recovery-rerun fix on BankLoanSta) is next at $-0.046$. PUFE and EMLI, which maintain two specialised predictors for the feature-absent and feature-present populations, come after and are essentially tied within seed noise at $-0.056$ and $-0.057$ respectively. OCDS, which jointly optimises a feature-relatedness graph and an $\ell_1$-truncated linear classifier, is the weakest ($-0.093$). Each gap is large compared to the seed-level variance observable in the per-cell std columns of Tables~\ref{tab:baseline_base_auc}--\ref{tab:baseline_ext_auc}, so the ordering is robust to seed under equal-tuning-budget terms (15 Optuna trials per (baseline, dataset, seed) cell, 5 seeds per cell). The $+0.039$ gap over GBDT-IL is the most informative one: GBDT-IL shares our learner (XGBoost) and an incremental warm-start mechanism, so what it lacks is the $\{$base, extended$\}$ model decomposition. The persistent gap therefore isolates the contribution of that architectural choice independently of the choice of learner. By contrast, the gap to the data-sharing Baseline Combined model is within sampling noise; the takeaway is that the decomposition closes the gap to a data-sharing model rather than surpassing it.

\subsection{Why Incremental Learning Succeeds Under Structured Missingness}

A natural question is why incremental learning without data sharing incurs no performance penalty. Our results identify two design mechanisms that explain how the lack of joint training is compensated:

\textbf{Effective knowledge transfer through warm-starting.} The extended model inherits learned patterns from the base model and refines them with newly available features rather than learning from scratch. Adaptive pruning determines how much base knowledge to retain, preventing negative transfer while preserving useful patterns. This is the mechanism that lets Phase~2 reach the operating point of a combined model trained on all rows with all features, while requiring only the model artefact (not the original training data) from Phase~1.

\textbf{Population specialisation as a byproduct of the incremental design.} The base model trains on all rows with extended features as NaN, so it implicitly specialises on the feature-absent subpopulation. This is an emergent property rather than an explicit design goal. The 5-seed picture in Table~\ref{tab:merged_auc_results} supports this view: the per-dataset gap (Our base $-$ BL base) is at most $\pm 0.014$ in mean across all 10 datasets and is consistent with sampling noise on every row. Modern GBMs handle absent features through heuristic default-direction routing; our seed-aware results are consistent with the view that dedicated incremental training is at least as good as that heuristic for the feature-absent subpopulation, which is what the parity claim requires.

The broader implication is operational: \textit{incremental learning without data sharing does not sacrifice performance in expectation}. The signed weighted-improvement column has 7 positive and 3 negative means across all 10 datasets, with paired Wilcoxon $p = 0.121$ (one-sided) over the $5\times 10 = 50$ seed-dataset pairs --- so we read the table as parity, not as a strict gain. Organisations can update models as new features arrive, avoid full retraining, and maintain data separation between phases without paying an accuracy price for those constraints. This makes the incremental approach a viable deployment choice when data sharing is constrained or when features evolve over time, even though we do not claim it strictly improves over a (data-sharing) combined model.

\subsection{Framework Contributions}

Despite the performance advantages, we emphasize that the main contribution of our work is the framework itself, which offers improved maintainability. The modular design allows for easier updates, monitoring, and extension compared to existing incremental learning solutions. Our framework also incorporates a critical rollback mechanism by concept, where in cases of significant data drift, it will rebuild the model from scratch rather than using the base model as input. This is accomplished by pruning all trees in the base model, effectively ensuring that outdated or irrelevant patterns don't negatively impact performance when data characteristics change substantially. This automatic adaptation mechanism provides an important safeguard for maintaining model reliability in dynamic environments.

\subsection{Application to Privacy-Constrained Model Collaboration}

Beyond feature evolution within a single organization, our experimental setup directly demonstrates a privacy-constrained collaboration scenario. In each experiment, the base model is trained on one feature set and the extended model incorporates additional features from a different subpopulation - no raw data is shared between the two phases, only the base model artifact. This mirrors a real-world deployment where Institution A trains a base model on its available features and shares the model file, while Institution B extends it with private features that Institution A never accesses.

All 10 datasets in our evaluation validate this capability: the two-phase approach achieves parity with a combined model that would require access to all features simultaneously --- per-dataset AUCs are within $\sim 1\%$ on every dataset and the average gap is indistinguishable from zero in our seed-variance check. This provides empirical evidence that model-level knowledge transfer can substitute for data-level sharing in cross-institutional settings such as healthcare (hospitals with different diagnostic panels) or finance (banks with complementary transaction features for AML detection).

We note that sharing model artifacts does not constitute formal privacy preservation - gradient boosting models can be susceptible to model inversion attacks~\cite{fredrikson2015model}. Production deployments under strict regulatory constraints (GDPR, HIPAA) would benefit from additional safeguards such as differential privacy or secure aggregation. Nonetheless, the framework reduces the attack surface significantly compared to raw data sharing, and the model-sharing paradigm represents a practical middle ground between full data exchange and complete isolation.

\subsection{Generalizability and Limitations}

While all datasets in our experiments focus on binary classification, the methodology naturally extends to regression and multi-class tasks. For regression, the same adaptive pruning mechanism applies with appropriate loss functions. For multi-class classification, the approach can be implemented via one-vs-rest decomposition or native multi-class boosting with minimal modification.

A more honest limitation concerns the gap between deployment and benchmarking. In deployment the base/extended allocation is given by the data sources --- the framework consumes it as input and runs Phases~1--2 directly, with no allocation search. Public benchmarks, however, do not expose such a source-derived split, so we used a benchmark-construction step (Section~\ref{sec:feature-allocation}) to screen candidate allocations restricted to NMAR-consistent groups and pick a representative one per dataset. This means our parity claim is conditional on a deployment whose natural allocation has the same NMAR structure as the screened benchmark allocations; we cannot rule out that deployments with a fundamentally different allocation regime would behave differently. With only 9-10 datasets to fit on, we likewise cannot identify a measurable dataset property that predicts \emph{a priori} which allocations will produce the largest parity-preserving gain, so a benchmark with a different screening protocol could land on different per-dataset numbers even though the framework code is unchanged. Additionally, run-to-run variance from the hyperparameter search process is substantial relative to the per-dataset weighted-improvement signal --- the same property that drives the parity (rather than strict-improvement) reading of Table~\ref{tab:merged_auc_results} --- so multiple seed runs are needed for stable per-dataset estimates.

\subsection{Future Work}

Several directions merit further investigation. First, extending the framework to support \textit{multiple sequential feature expansions} (Phase 3, Phase 4, etc.) would better reflect real-world scenarios where features arrive in multiple waves over time. Second, developing theoretical guarantees for when incremental learning provably maintains parity with joint training would reduce reliance on empirical validation. Third, exploring whether information-theoretic properties of the extended features (e.g., conditional mutual information with the target given base features) can predict incremental learning success would enable more principled feature allocation without exhaustive search. Finally, integrating our approach with privacy-preserving techniques such as differential privacy or secure aggregation could formalize the model-sharing paradigm described above.

\section{Conclusion}

We presented an incremental learning framework for Gradient Boosting Machines that integrates new features into existing models without full retraining. Our two-phase approach trains a base model on commonly available features and incrementally extends it with newly available features through warm-starting and adaptive tree pruning - with only the model artifact shared between phases, enabling both incremental updates and privacy-constrained deployment.

Evaluation across 10 diverse datasets from healthcare, finance, and meteorology domains demonstrates that incremental learning without data sharing matches standard training: per-dataset AUCs sit within $\sim 1\%$ of a data-sharing combined model on every dataset, and the population-weighted gap between the two approaches is indistinguishable from zero in expectation across train/test splits. This is the central result: \textit{data separation between training phases incurs no performance cost} - while providing operational benefits that standard training cannot offer, including incremental updates and privacy-constrained deployment.

A head-to-head comparison against five recent baselines for learning with evolving feature spaces - AdaptiveXGBoost~\cite{montiel2020axgb}, PUFE~\cite{he2021pufe}, OCDS~\cite{he2019ocds}, EMLI~\cite{dong2021eml}, and GBDT-IL~\cite{chen2024gbdtil} - run under an identical preprocessing and data-split protocol shows that our framework's mean population-weighted AUC across the 10 datasets is $+0.039$ above the closest competitor (GBDT-IL, which shares our GBDT backbone) and up to $+0.09$ above the weakest (OCDS), under an equal-tuning-budget protocol (15 Optuna trials per baseline-dataset-seed cell, 5 seeds). These margins are an order of magnitude larger than the seed-level variance observable in the per-cell std columns of Tables~\ref{tab:baseline_base_auc}--\ref{tab:baseline_ext_auc} and therefore robust, isolating the contribution of the $\{$base, extended$\}$ model decomposition beyond the choice of learner.

The adaptive pruning mechanism --- which determines how much prior knowledge to retain during incremental extension --- lands at the top of the per-dataset ordering on most datasets in our experiments, and by virtue of the $k=0$ branch of the search it cannot underperform training from scratch in expectation. The seed-aware Table~\ref{tab:merged_auc_results} (5 seeds per dataset, 10 datasets) highlights one mechanism by which no performance is lost despite data separation: the two-phase design produces population-specific specialisation as an emergent benefit, with the per-dataset gap between our base model and the combined baseline on the feature-absent subpopulation bounded by $\pm 0.014$ in mean and consistent with sampling noise on every row.

The practical implications are twofold. For \textit{evolving feature spaces}, the framework enables incremental model updates as new features become available, without requiring access to historical training data or full retraining. For \textit{privacy-constrained collaboration}, institutions can share model artifacts rather than raw data --- our 5-seed evaluation across all 10 datasets is consistent with model-level knowledge transfer achieving parity with full-data training (paired Wilcoxon $p = 0.121$ over the $5\times 10 = 50$ paired seed-dataset cells). Together, these capabilities address a growing need in healthcare, finance, and other regulated domains where both data evolution and data privacy are first-order concerns.

\section*{Supplementary Materials and Code Availability}

To support reproducibility and facilitate further research, we provide supplementary materials that include:

\begin{itemize}
  \item The full implementation of our two-phase incremental GBM framework.
  \item Experiments code, logs and hyperparameter configurations for all datasets.
  \item Sensitivity analysis.
  \item Ablation study.
  \item Link: https://anonymous.4open.science/r/Learning-to-Grow-Incremental-Trees-over-Expanding-Features-818A/
\end{itemize}


\section*{Generative AI Usage Disclosure}

In accordance with ACM's Policy on Authorship and the guidelines for disclosure of generative AI usage, the authors hereby declare the following:


\noindent\textbullet\ \textbf{Language editing.} We used OpenAI ChatGPT and Anthropic Claude solely to improve spelling, grammar, punctuation, clarity, and overall readability of the manuscript. All suggested edits were critically reviewed, revised as necessary, and confirmed by the authors to ensure that the text remains an accurate and faithful representation of our original intellectual contributions.\\[0.5em]
\noindent\textbullet\ \textbf{Code development.} We employed OpenAI ChatGPT and Anthropic Claude to help scaffold boilerplate and to accelerate routine implementation tasks. Every line of code generated or suggested by these tools was examined, adapted, and tested by the authors to guarantee that it performs exactly as intended.\\[0.5em]

\noindent The authors take full responsibility for the veracity and correctness of all material in this Work, and no generative AI system is listed as an author.


\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
